\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-09-26 challenge.tex
Challenge#2

Supporting inheritance and family polymorphism in a proof assistant poses challenges
far beyond those encountered in an object-oriented setting~\cite{ncm2004,vc-calculus-2006,zm2017},
because the underlying programming language is simultaneously
functional, dependently typed, a logic, and an interactive tool.

@@

A family-polymorphism mechanism could potentially support extensible
inductive types, by allowing a \emph{derived family} to add new constructors
to inductive types inherited from a \emph{base family}.
Such capability would be useful for extending mechanized languages.
For example, \cref{fig:?} shows the part of an STLC extension that could
potentially be mechanized using such extensible inductive types: a derived family
inherits constructors of inductive types from the base family (viz.\ the $\cdots$'s)
and adds new constructors to model the new language feature.

@@

However, there is a tension between extensibility of inductive types
and exhaustivity of induction reasoning.
In \cref{fig:stlc-nonmechanized}, all the lemmas and theorems, as well as
the substitution function, require induction reasoning.
In the derived family, to ensure that induction remains exhaustive,
how could the programmer account for the new cases arising from the new
constructors, without redefining or forcing the rechecking of those cases
already handled by the base family?

The tension results from the duality between variants and records.
For record-like language constructs (e.g., classes and objects), it is
safe for an extension to add more fields.
By contrast, variant-like language constructs (e.g., inductive types)
do not automatically enjoy safe, modular addition of constructors.

\noindentparagraph{Late binding vs.\ dependent typing.}

\noindentparagraph{Essence of inheritance vs.\ consistency of a logic.}

>>>>

Integrating family polymorphism into an interactive theorem prover poses challenges
far beyond those found in an object-oriented setting~\cite{ncm2004,vc-calculus-2006,zm2017},
because the underlying programming language is simultaneously
functional, dependently typed, a logic, and an interactive tool.

@@

A family-polymorphism mechanism could potentially support extensible
inductive types, by allowing a \emph{derived family} to add new constructors
to inductive types inherited from a \emph{base family}.
Such capability would be useful for extending mechanized languages.
For example, \cref{fig:stlc-nonmechanized} shows the part of an STLC extension that could
potentially be mechanized using such extensible inductive types: a derived family
inherits constructors of inductive types from the base family (viz.\ the ellipsis under ``Typing rule'')
and adds new constructors to model the new language feature.

@@

However, there is a tension between extensibility of inductive types
and exhaustivity of induction reasoning.
In \cref{fig:stlc-nonmechanized}, all the lemmas and theorems, as well as
the substitution function, require induction reasoning.
A language design must enforce that induction remains exhaustive in the
derived family despite the new constructors, without requiring
redefinition or rechecking of those cases already handled by the base
family.
%
The tension reflects a duality between variants and records.
For record-like language constructs (e.g., classes and objects), it is
safe for an extension to add more fields: projection of existent fields
still works.
But variant-like constructs (e.g., inductive types) do not automatically
enjoy safe, modular addition of constructors:
pattern matching could potentially become non-exhaustive.

\noindentparagraph{Overriding vs.\ dependent typing.}

Common to many family-polymorphism mechanisms is the ability for a
derived family to override fields defined in the base family.
As we show later, this expressiveness is also useful for metatheory
mechanization.
However, it is in conflict with dependent typing, wherein type checking
often involves unfolding definitions.
In \cref{fig:stlc-nonmechanized}, the proof of the substitution lemma is
checked against the definition of the substitution function.
If a derived family were allowed to arbitrarily override how
substitution is defined, then the proof, when inherited, would be
ill-formed.
A language design should only allow those overridings that are safe.

\noindentparagraph{Self reference vs.\ logical consistency.}

### 2022-09-26 introduction.tex
Challenge#2

In the Coq proof assistant~\cite{coq}, inductive types, as well as functions and
theorems over inductive types, are closed to extension.
So to reuse mechanized metatheories,
the common practice is still to copy code and proofs and then modify them in each extension.
But having to maintain multiple copies is highly non-modular and
antithetical to good software engineering.
%
The programmer could also turn to design patterns~\cite{delaware2011,delaware2013}.
But they tend to require heavy lifting from the programmer to make code
extensible, often leading to verbose, non-idiomatic programming styles.

We answer the following question:
can linguistic features designed to solve the EP (in
conventional OO and functional languages)
be adapted to metatheory mechanization in a proof assistant?

@@

Language mechanisms including
mixins~\cite{mixin-1990},
virtual classes~\cite{virtualclasses-1989,vc-calculus-2006},
virtual types~\cite{thorup97} and associated types~\cite{ckj05},
extensible cases~\cite{bac2006},
and so on, are all based on this essential idea of inheritance.
%
In particular, when a language mechanism allows inheritance over
recursive definitions containing related types and terms,
it is said to support \emph{family polymorphism}~\cite{ernst2001family}:
types and terms are polymorphic to a family they are nested within.

@@

%\noindent
%As an example, consider \cref{fig:STLC-example}, the simply typed
%lambda calculus (STLC) mechanized in Coq.
%A programming challenge there is to define distinct extensions of this
%STLC formalization to support distinct features,
%and then selectively compose these extensions to form new STLC variants.
As an example, the diagram to the right depicts an extensible
mechanization of the simply typed λ-calculus (STLC), using family
polymorphism.
An extension of STLC with products and another with iso-recursive types
can both inherit from the base STLC family:
they reuse mechanized metatheories,
from abstract syntax all the way to the type-safety theorem,
only adding new constructors to inductive types
and adding new cases to recursive functions and induction proofs
\emph{as needed} by an extension.

>>>>

In the Coq proof assistant~\cite{coq}, for instance, inductive types, as well as functions and
theorems over inductive types, are closed to extension.
So to reuse mechanized metatheories,
the common practice is still to copy code and proofs and then modify them in each extension.
But having to maintain multiple copies is highly non-modular and
antithetical to good software engineering.
%
The programmer could also turn to design patterns~\cite{delaware2011,delaware2013} that
use software product lines or Church encodings.
But they tend to require heavy lifting from the programmer to make code
extensible, often leading to non-idiomatic programming styles.

We thus set out to answer the following question:
can extensible metatheory mechanization be made easier by
having a proof assistant support new \emph{linguistic features} that address the EP?
%can linguistic features designed to solve the EP for designed to solve the EP (in
%conventional OO and functional languages)
%be adapted to metatheory mechanization in a proof assistant?

@@

Language mechanisms including
mixins~\cite{mixin-1990},
virtual classes~\cite{virtualclasses-1989,vc-calculus-2006},
virtual types~\cite{thorup97}, % and associated types~\cite{ckj05},
extensible cases~\cite{bac2006}, etc.\ 
are all based on this essential idea of inheritance.
%
In particular, when a language mechanism allows inheritance over
recursive definitions containing related types and terms,
it is said to support \emph{family polymorphism}~\cite{ernst2001family}:
types and terms are polymorphic to a family they are nested within.

@@

%\noindent
%As an example, consider \cref{fig:STLC-example}, the simply typed
%lambda calculus (STLC) mechanized in Coq.
%A programming challenge there is to define distinct extensions of this
%STLC formalization to support distinct features,
%and then selectively compose these extensions to form new STLC variants.
As an example, the diagram to the right depicts an extensible
mechanization of the simply typed λ-calculus (STLC), using family
polymorphism.
An extension of STLC with products and another with fixpoints
can both inherit from the base STLC family:
they reuse mechanized metatheories,
from abstract syntax all the way to the type-safety theorem,
only adding new constructors to inductive types
and adding new cases to recursive functions and induction proofs
\emph{as needed} by an extension.
